---
title: 'Getting started; Get the R scripts'
author: "Marc A.T. Teunis"
date: "`r Sys.Date()`"
output:
  ioslides_presentation: 
    highlight: pygments
    widescreen: yes
runtime: shiny_prerendered
---
<style>

.footer {
    font-size: 16pt;
    color: Black;
    font-family: 'Helvetica'
    content: attr(label);
    position: absolute;
    margin-left: auto;
    margin-right: auto;
    left: 0;
    right: 0;
    text-align:center;
    bottom: 10px;
    line-height: 1.9;
    display: block;
}

</style>

<div class="footer" style="margin-top;font-size:120%;"> 
Data Analysis Using R - November 2017 </div>
```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, fig.width = 5, fig.height = 3)
```

```{r, root_1, include=FALSE}
## defines the root of the project for later use
require("rprojroot") || utils::install.packages("rprojroot")
library(rprojroot)
root <- find_root_file(criterion = is_rstudio_project)
```

## The R scripts

The book "Discovering Statitics Using R" has many resources online. The R scripts used in the book can be downloaded as zip archives. In the code below we extract these archives so that we can work with them in R.
The zip files are located in the folder `./code`. After running the code below the R code from the book chapters will be available in the folder `./code`.

```{r}
path_to_zipfiles <- file.path(root, "code")
zipF <- list.files(path = path_to_zipfiles, pattern=".zip", full.names = T); sapply(zipF, unzip, exdir = path_to_zipfiles)

```


