---
title: "Several R plot examples"
author: "Marc A.T. Teunis"
date: "`r Sys.Date()`"
output:
  ioslides_presentation: 
    highlight: pygments
    widescreen: yes
---
<style>

.footer {
    font-size: 16pt;
    color: Black;
    font-family: 'Helvetica'
    content: attr(label);
    position: absolute;
    margin-left: auto;
    margin-right: auto;
    left: 0;
    right: 0;
    text-align:center;
    bottom: 10px;
    line-height: 1.9;
    display: block;
}

</style>

<div class="footer" style="margin-top;font-size:120%;"> 
Data Analysis Using R - January 2018 </div>

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE,
                      warning = FALSE,
                      error = FALSE,
                      message = FALSE)

```

## In the intro lecture we saw the CIRCOS plot.
The code of this plot is depicted in the `code chunk` below.
```{r}
# Press the `knit` button in RStudio and look at the result.
```

## R packages
```{r, echo=TRUE}
## CRAN
require("stats") || utils::install.packages("stats")
library(stats)
require("circlize") || utils::install.packages("circlize")
library(circlize) 
library(pheatmap)
library(ggplot2)
library(drc)
library(tidyverse)
## Bionconductor
# source("https://bioconductor.org/biocLite.R")
# biocLite("DESeq2")
#biocLite("GEOquery")
#biocLite("tximport")
#biocLite("tximportData")
biocLite("pasilla")
library(DESeq2)
library(GEOquery)
library("tximport")
library("readr")
library("tximportData")
library(pasilla)
```

## Circos plot
```{r}
# dend: as dendrogram object, usually returned by hclust and as.dendrogram
# maxy: maximum height of the tree
circos.dendrogram = function(dend, maxy) {
  labels = as.character(labels(dend))
  x = seq_along(labels) - 0.5
  names(x) = labels
  
  is.leaf = function(object) (is.logical(L <- attr(object, "leaf"))) && L
  
  # recursive function to draw the tree
  draw.d = function(dend, maxy) {
    leaf = attr(dend, "leaf")
    d1 = dend[[1]]
    d2 = dend[[2]]
    height = attr(dend, 'height')
    midpoint = attr(dend, 'midpoint')
    
    if(is.leaf(d1)) {
      x1 = x[as.character(attr(d1, "label"))]
    } else {
      x1 = attr(d1, "midpoint") + x[as.character(labels(d1))[1]]
    }
    y1 = attr(d1, "height")
    
    if(is.leaf(d2)) {
      x2 = x[as.character(attr(d2, "label"))]
    } else {
      x2 = attr(d2, "midpoint") + x[as.character(labels(d2))[1]]
    }
    y2 = attr(d2, "height")
    
    circos.lines(c(x1, x1), maxy - c(y1, height), straight = TRUE)
    circos.lines(c(x1, x2), maxy - c(height, height))
    circos.lines(c(x2, x2), maxy - c(y2, height), straight = TRUE)
    
    if(!is.leaf(d1)) {
      draw.d(d1, maxy)
    }
    if(!is.leaf(d2)) {
      draw.d(d2, maxy)
    }
  }
  
  draw.d(dend, maxy)
}


mat = matrix(rnorm(100*10), nrow = 10, ncol = 100)
factors = rep(letters[1:2], 50)
par(mar = c(1, 1, 1, 1))
circos.par(cell.padding = c(0, 0, 0, 0), gap.degree = 5)
circos.initialize(factors, xlim = c(0, 50))
maxy = 0

f = colorRamp2(breaks = c(-1, 0, 1), colors = c("green", "black", "red"))

circos.trackPlotRegion(ylim = c(0, 10), bg.border = NA, panel.fun = function(x, y) {
  sector.index = get.cell.meta.data("sector.index")
  m = mat[, factors == sector.index]
  
  dend.col = as.dendrogram(hclust(dist(t(m))))
  
  maxy = ifelse(maxy > attr(dend.col, "height"), maxy, attr(dend.col, "height"))
  assign("maxy", maxy, envir = .GlobalEnv)
  
  m2 = m[, labels(dend.col)]
  nr = nrow(m2)
  nc = ncol(m2)
  for(i in 1:nr) {
    for(j in 1:nc) {
      circos.rect(j-1, nr-i, j, nr-i+1, border = f(m2[i, j]), col = f(m2[i, j]))
    }
  }
  
})
circos.trackPlotRegion(ylim = c(0, maxy), bg.border = NA, track.height = 0.3, panel.fun = function(x, y) {
  sector.index = get.cell.meta.data("sector.index")
  m = mat[, factors == sector.index]
  
  dend.col = as.dendrogram(hclust(dist(t(m))))
  
  circos.dendrogram(dend.col, maxy)
  
})

circos.clear()

x = seq(-10, 10, length.out=100)/40
col =f(seq(-2, 2, length.out = length(x-1)))
for(i in seq_along(x)) {
  if(i == 1) next
  rect(x[i-1], -0.05, x[i], 0.05, col = col[i], border = col[i])
}

text(x[1], -0.08, "-2", adj = c(0.5, 1), cex = 1.2)
text(x[ceiling(length(x)/2)], -0.08, "0", adj = c(0.5, 1), cex = 1.2)
text(x[length(x)], -0.08, "2", adj = c(0.5, 1), cex = 1.2)
```

## Heatmaps
```{r}
library(gplots)
library(geneplotter)

# Mol Biol Cell. 1998 Dec;9(12):3273-97.
# Comprehensive identification of cell cycle-regulated genes of the yeast
# Saccharomyces cerevisiae by microarray hybridization.
# Spellman PT, Sherlock G, Zhang MQ, Iyer VR, Anders K, Eisen MB, Brown PO, Botstein D, Futcher B.
# Department of Genetics, Stanford University Medical Center, Stanford, California 94306-5120, USA.
#
# We sought to create a comprehensive catalog of yeast genes whose transcript levels vary
# periodically within the cell cycle. To this end, we used DNA microarrays and samples from
# yeast cultures synchronized by three independent methods: alpha factor arrest, elutriation,
# and arrest of a cdc15 temperature-sensitive mutant. Using periodicity and correlation algorithms,
# we identified 800 genes that meet an objective minimum criterion for cell cycle regulation.
# In separate experiments, designed to examine the effects of inducing either the G1 cyclin
# Cln3p or the B-type cyclin Clb2p, we found that the mRNA levels of more than half of these
# 800 genes respond to one or both of these cyclins. Furthermore, we analyzed our set of cell
# cycle-regulated genes for known and new promoter elements and show that several known elements
# (or variations thereof) contain information predictive of cell cycle regulation. A full
# description and complete data sets are available at http://cellcycle-www.stanford.edu

# read the data in from URL
bots <- read.table(url("http://genome-www.stanford.edu/cellcycle/data/rawdata/combined.txt"), sep="\t", header=TRUE)

# get just the alpha data
abot <- bots[,c(8:25)]
rownames(abot) <- bots[,1]
abot[1:7,] %>% as_tibble()

# get rid of NAs
abot[is.na(abot)] <- 0

# we need to find a way of reducing the data. Can't do ANOVA as there are no
# replicates. Sort on max difference and take first 1000
min <-apply(abot, 1, min)
max <- apply(abot, 1, max)
sabot <- abot[order(max - min, decreasing=TRUE),][1:1000,]

# cluster on correlation
cdist <- as.dist(1 - cor(t(sabot)))
hc <- hclust(cdist, "average")
```

## Draw a heatmap
```{r}
# x11()
heatmap.2(as.matrix(sabot),
          Rowv=as.dendrogram(hc),
          Colv=FALSE,
          cexRow=0.05,
          cexCol=1,
          dendrogram="row",
          scale="row",
          trace="none",
          density.info="none",
          key=FALSE,
          col=greenred.colors(80))
```

## Different colours
```{r}
# x11()
heatmap.2(as.matrix(sabot),
          Rowv=as.dendrogram(hc),
          Colv=FALSE,
          cexRow=0.05,
          cexCol=1,
          dendrogram="row",
          scale="row",
          trace="none",
          density.info="none",
          key=FALSE,
          col=heat.colors(80))
```

## More control over colours
```{r}
# x11()

my_palette <- colorRampPalette(c("blue", "yellow", "darkgreen"))(n = 80)

heatmap.2(as.matrix(sabot),
          Rowv=as.dendrogram(hc),
          Colv=FALSE,
          cexRow=0.05,
          cexCol=1,
          dendrogram="row",
          scale="row",
          trace="none",
          density.info="none",
          key=FALSE,
          col=my_palette)
```

## Dose response relations
```{r, echo=TRUE}
library(drc)
ryegrass %>% as_tibble()
## Fitting models to be plotted below
ryegrass.m1 <- drm(rootl~conc, data = ryegrass, fct = LL.4())
ryegrass.m2 <- drm(rootl~conc, data = ryegrass, fct = LL.3())  
# lower limit fixed at 0
```

## Plotting observations and fitted curve for the first model
```{r}
library(drc)
plot(ryegrass.m1, broken = TRUE)
```

## Adding fitted curve for the second model (not much difference)
```{r}
plot(ryegrass.m1, broken = TRUE)
plot(ryegrass.m2, broken = TRUE, add = TRUE, type = "none", col = 2, lty = 2)
```

## Add confidence region for the first model.
```{r}
plot(ryegrass.m1, broken = TRUE)
plot(ryegrass.m2, broken = TRUE, add = TRUE, type = "none", col = 2, lty = 2)
plot(ryegrass.m1, broken = TRUE, type="confidence", add=TRUE) 
```

## Finetuning the axis break
```{r}
plot(ryegrass.m1, broken = TRUE, bcontrol = list(style = "gap"))
```

## Network plots
```{r}
library(ReactomePA)
## try http:// if https:// URLs are not supported
source("https://bioconductor.org/biocLite.R")
#biocLite("ReactomePA")
#biocLite("DOSE")
#biocLite("org.Hs.eg.db")
#browseVignettes("ReactomePA")

library(org.Hs.eg.db)
library(DOSE)
library(ReactomePA)

## ------------------------------------------------------------------------
library(ReactomePA)
data(geneList)
de <- names(geneList)[abs(geneList) > 1.5]
head(de)
x <- enrichPathway(gene=de,pvalueCutoff=0.05, readable=T)
head(as.data.frame(x))
```

## Barplot
```{r, fig.height=6, fig.width=12}
## ----fig.height=6, fig.width=12------------------------------------------

barplot(x, showCategory=8)
```

## Dotplot
```{r}
## ----fig.height=6, fig.width=12------------------------------------------
dotplot(x, showCategory=15)
```

## Network with clusters
```{r}
## ----fig.height=10, fig.width=10-----------------------------------------
emapplot(x)
```

## Cnet network
```{r}
## ----fig.height=8, fig.width=8-------------------------------------------
cnetplot(x, categorySize="pvalue", foldChange=geneList)
```

## ggplot2
```{r}
library(ggplot2)
library(dplyr)
library(gapminder)
library(wesanderson)

gap_07 <- filter(gapminder, year == 2007)

ggplot(gap_07, aes(x = gdpPercap, y = lifeExp,
                   color = continent)) +

  geom_point(size = 2) +
  scale_x_log10() +
  theme_light() +
  theme(legend.position = c(0.1, 0.85),
legend.key = element_blank(),
axis.text = element_text(size = 12),
axis.title = element_text(size = 14)) +
  labs(x = "Per capita GDP",
       y = "Life Expectancy",
       title = "2007 Life Expectancy and GDP",
        color = "Continent") +
  ggsave("life_exp_gdp_2007_cutom.png", 
width = 7, height = 7)
```

## Wells plates
```{r}
# install.packages("platetools")
library(platetools)
library(viridis)

# example dataframe
df <- data.frame(vals = rnorm(384),
                 well = num_to_well(1:384, plate = 384))

raw_map(data = df$vals,
        well = df$well,
        plate = 384) +
    ggtitle("Example 384-well plate") +
    theme_dark() +
    scale_fill_viridis()
```

## RNA sequencing
```{r}
library(DESeq2)
#browseVignettes('DESeq2')
## ----setup, echo=FALSE, results="hide"-----------------------------------

## ----quickStart, eval=FALSE----------------------------------------------
#  dds <- DESeqDataSetFromMatrix(countData = cts,
 #                               colData = coldata,
  #                              design= ~ batch + condition)
#  dds <- DESeq(dds)
#  resultsNames(dds) # lists the coefficients
#  res <- results(dds, name="condition_trt_vs_untrt")
#  # or to shrink log fold changes association with condition:
#  res <- lfcShrink(dds, coef="condition_trt_vs_untrt", type="apeglm")

## ----txiSetup------------------------------------------------------------

dir <- system.file("extdata", package="tximportData")
samples <- read.table(file.path(dir,"samples.txt"), header=TRUE)
samples$condition <- factor(rep(c("A","B"),each=3))
rownames(samples) <- samples$run
samples[,c("pop","center","run","condition")]

## ----txiFiles------------------------------------------------------------
files <- file.path(dir,"salmon", samples$run, "quant.sf.gz")
names(files) <- samples$run
tx2gene <- read_csv(file.path(dir, "tx2gene.gencode.v27.csv"))

## ----tximport, results="hide"--------------------------------------------
txi <- tximport(files, type="salmon", tx2gene=tx2gene)

## ----txi2dds, results="hide"---------------------------------------------
library("DESeq2")
ddsTxi <- DESeqDataSetFromTximport(txi,
                                   colData = samples,
                                   design = ~ condition)

## ----loadPasilla---------------------------------------------------------
library("pasilla")
pasCts <- system.file("extdata",
                      "pasilla_gene_counts.tsv",
                      package="pasilla", mustWork=TRUE)
pasAnno <- system.file("extdata",
                       "pasilla_sample_annotation.csv",
                       package="pasilla", mustWork=TRUE)
cts <- as.matrix(read.csv(pasCts,sep="\t",row.names="gene_id"))
coldata <- read.csv(pasAnno, row.names=1)
coldata <- coldata[,c("condition","type")]

## ----showPasilla---------------------------------------------------------
head(cts,2)
coldata

## ----reorderPasila-------------------------------------------------------
rownames(coldata) <- sub("fb", "", rownames(coldata))
all(rownames(coldata) %in% colnames(cts))
all(rownames(coldata) == colnames(cts))
cts <- cts[, rownames(coldata)]
all(rownames(coldata) == colnames(cts))

## ----matrixInput---------------------------------------------------------
library("DESeq2")
dds <- DESeqDataSetFromMatrix(countData = cts,
                              colData = coldata,
                              design = ~ condition)
dds

## ----addFeatureData------------------------------------------------------
featureData <- data.frame(gene=rownames(cts))
mcols(dds) <- DataFrame(mcols(dds), featureData)
mcols(dds)

## ----htseqDirI, eval=FALSE-----------------------------------------------
#  directory <- "/path/to/your/files/"

## ----htseqDirII----------------------------------------------------------
directory <- system.file("extdata", package="pasilla",
                         mustWork=TRUE)

## ----htseqInput----------------------------------------------------------
sampleFiles <- grep("treated",list.files(directory),value=TRUE)
sampleCondition <- sub("(.*treated).*","\\1",sampleFiles)
sampleTable <- data.frame(sampleName = sampleFiles,
                          fileName = sampleFiles,
                          condition = sampleCondition)

## ----hsteqDds------------------------------------------------------------
library("DESeq2")
ddsHTSeq <- DESeqDataSetFromHTSeqCount(sampleTable = sampleTable,
                                       directory = directory,
                                       design= ~ condition)
ddsHTSeq

## ----loadSumExp----------------------------------------------------------
# biocLite("airway")
library("airway")
data("airway")
se <- airway

## ----sumExpInput---------------------------------------------------------
library("DESeq2")
ddsSE <- DESeqDataSet(se, design = ~ cell + dex)
ddsSE

## ----prefilter-----------------------------------------------------------
keep <- rowSums(counts(dds)) >= 10
dds <- dds[keep,]

## ----factorlvl-----------------------------------------------------------
dds$condition <- factor(dds$condition, levels = c("untreated","treated"))

## ----relevel-------------------------------------------------------------
dds$condition <- relevel(dds$condition, ref = "untreated")

## ----droplevels----------------------------------------------------------
dds$condition <- droplevels(dds$condition)

## ----deseq---------------------------------------------------------------
dds <- DESeq(dds)
res <- results(dds)
res

## ----eval=FALSE----------------------------------------------------------
#  res <- results(dds, name="condition_treated_vs_untreated")
#  res <- results(dds, contrast=c("condition","treated","untreated"))

## ----lfcShrink-----------------------------------------------------------
#biocLite("apeglm")
resultsNames(dds)
resLFC <- lfcShrink(dds, coef="condition_treated_vs_untreated", type="apeglm")
resLFC

## ----parallel, eval=FALSE------------------------------------------------
#  library("BiocParallel")
#  register(MulticoreParam(4))

## ----resOrder------------------------------------------------------------
resOrdered <- res[order(res$pvalue),]

## ----sumRes--------------------------------------------------------------
summary(res)

## ----sumRes01------------------------------------------------------------
sum(res$padj < 0.1, na.rm=TRUE)

## ----resAlpha05----------------------------------------------------------
res05 <- results(dds, alpha=0.05)
summary(res05)
sum(res05$padj < 0.05, na.rm=TRUE)

## ----IHW-----------------------------------------------------------------
#biocLite("IHW")
library("IHW")
resIHW <- results(dds, filterFun=ihw)
summary(resIHW)
sum(resIHW$padj < 0.1, na.rm=TRUE)
metadata(resIHW)$ihwResult

## ----MA------------------------------------------------------------------
plotMA(res, ylim=c(-2,2))

## ----shrunkMA------------------------------------------------------------
plotMA(resLFC, ylim=c(-2,2))

## ----MAidentify, eval=FALSE----------------------------------------------
#  idx <- identify(res$baseMean, res$log2FoldChange)
#  rownames(res)[idx]

## ----warning=FALSE-------------------------------------------------------
resultsNames(dds)
# because we are interested in treated vs untreated, we set 'coef=2'
resNorm <- lfcShrink(dds, coef=2, type="normal")
resAsh <- lfcShrink(dds, coef=2, type="ashr")

biocLite("ashr")
## ----fig.width=8, fig.height=3-------------------------------------------
par(mfrow=c(1,3), mar=c(4,4,2,1))
xlim <- c(1,1e5); ylim <- c(-3,3)
plotMA(resLFC, xlim=xlim, ylim=ylim, main="apeglm")
plotMA(resNorm, xlim=xlim, ylim=ylim, main="normal")
plotMA(resAsh, xlim=xlim, ylim=ylim, main="ashr")

## ----plotCounts----------------------------------------------------------
plotCounts(dds, gene=which.min(res$padj), intgroup="condition")

## ----plotCountsAdv-------------------------------------------------------
d <- plotCounts(dds, gene=which.min(res$padj), intgroup="condition", 
                returnData=TRUE)
library("ggplot2")
ggplot(d, aes(x=condition, y=count)) + 
  geom_point(position=position_jitter(w=0.1,h=0)) + 
  scale_y_log10(breaks=c(25,100,400))

## ----metadata------------------------------------------------------------
mcols(res)$description

## ----export, eval=FALSE--------------------------------------------------
#  write.csv(as.data.frame(resOrdered),
#            file="condition_treated_results.csv")

## ----subset--------------------------------------------------------------
resSig <- subset(resOrdered, padj < 0.1)
resSig

## ----multifactor---------------------------------------------------------
colData(dds)

## ----copyMultifactor-----------------------------------------------------
ddsMF <- dds

## ----fixLevels-----------------------------------------------------------
levels(ddsMF$type)
levels(ddsMF$type) <- sub("-.*", "", levels(ddsMF$type))
levels(ddsMF$type)

## ----replaceDesign-------------------------------------------------------
design(ddsMF) <- formula(~ type + condition)
ddsMF <- DESeq(ddsMF)

## ----multiResults--------------------------------------------------------
resMF <- results(ddsMF)
head(resMF)

## ----multiTypeResults----------------------------------------------------
resMFType <- results(ddsMF,
                     contrast=c("type", "single", "paired"))
head(resMFType)

## ----rlogAndVST----------------------------------------------------------
vsd <- vst(dds, blind=FALSE)
rld <- rlog(dds, blind=FALSE)
head(assay(vsd), 3)

## ----meansd--------------------------------------------------------------
# this gives log2(n + 1)
ntd <- normTransform(dds)
library("vsn")
meanSdPlot(assay(ntd))
meanSdPlot(assay(vsd))
meanSdPlot(assay(rld))

## ----heatmap-------------------------------------------------------------
library("pheatmap")
select <- order(rowMeans(counts(dds,normalized=TRUE)),
                decreasing=TRUE)[1:20]
df <- as.data.frame(colData(dds)[,c("condition","type")])
pheatmap(assay(ntd)[select,], cluster_rows=FALSE, show_rownames=FALSE,
         cluster_cols=FALSE, annotation_col=df)
pheatmap(assay(vsd)[select,], cluster_rows=FALSE, show_rownames=FALSE,
         cluster_cols=FALSE, annotation_col=df)
pheatmap(assay(rld)[select,], cluster_rows=FALSE, show_rownames=FALSE,
         cluster_cols=FALSE, annotation_col=df)

## ----sampleClust---------------------------------------------------------
sampleDists <- dist(t(assay(vsd)))

## ----figHeatmapSamples, fig.height=4, fig.width=6------------------------
library("RColorBrewer")
sampleDistMatrix <- as.matrix(sampleDists)
rownames(sampleDistMatrix) <- paste(vsd$condition, vsd$type, sep="-")
colnames(sampleDistMatrix) <- NULL
colors <- colorRampPalette( rev(brewer.pal(9, "Blues")) )(255)
pheatmap(sampleDistMatrix,
         clustering_distance_rows=sampleDists,
         clustering_distance_cols=sampleDists,
         col=colors)

## ----figPCA--------------------------------------------------------------
plotPCA(vsd, intgroup=c("condition", "type"))

## ----figPCA2-------------------------------------------------------------
pcaData <- plotPCA(vsd, intgroup=c("condition", "type"), returnData=TRUE)
percentVar <- round(100 * attr(pcaData, "percentVar"))
ggplot(pcaData, aes(PC1, PC2, color=condition, shape=type)) +
  geom_point(size=3) +
  xlab(paste0("PC1: ",percentVar[1],"% variance")) +
  ylab(paste0("PC2: ",percentVar[2],"% variance")) + 
  coord_fixed()

## ----WaldTest, eval=FALSE------------------------------------------------
#  dds <- estimateSizeFactors(dds)
#  dds <- estimateDispersions(dds)
#  dds <- nbinomWaldTest(dds)

## ----simpleContrast, eval=FALSE------------------------------------------
#  results(dds, contrast=c("condition","C","B"))

## ----combineFactors, eval=FALSE------------------------------------------
#  dds$group <- factor(paste0(dds$genotype, dds$condition))
#  design(dds) <- ~ group
#  dds <- DESeq(dds)
#  resultsNames(dds)
#  results(dds, contrast=c("group", "IB", "IA"))

## ----interFig, echo=FALSE, results="hide", fig.height=3------------------
npg <- 20
mu <- 2^c(8,10,9,11,10,12)
cond <- rep(rep(c("A","B"),each=npg),3)
geno <- rep(c("I","II","III"),each=2*npg)
table(cond, geno)
counts <- rnbinom(6*npg, mu=rep(mu,each=npg), size=1/.01)
d <- data.frame(log2c=log2(counts+1), cond, geno)
library("ggplot2")
plotit <- function(d, title) {
  ggplot(d, aes(x=cond, y=log2c, group=geno)) + 
    geom_jitter(size=1.5, position = position_jitter(width=.15)) +
    facet_wrap(~ geno) + 
    stat_summary(fun.y=mean, geom="line", colour="red", size=0.8) + 
    xlab("condition") + ylab("log2(counts+1)") + ggtitle(title)
}
plotit(d, "Gene 1") + ylim(7,13)
lm(log2c ~ cond + geno + geno:cond, data=d)

## ----interFig2, echo=FALSE, results="hide", fig.height=3-----------------
mu[4] <- 2^12
mu[6] <- 2^8
counts <- rnbinom(6*npg, mu=rep(mu,each=npg), size=1/.01)
d2 <- data.frame(log2c=log2(counts + 1), cond, geno)
plotit(d2, "Gene 2") + ylim(7,13)
lm(log2c ~ cond + geno + geno:cond, data=d2)

## ----simpleLRT, eval=FALSE-----------------------------------------------
#  dds <- DESeq(dds, test="LRT", reduced=~1)
#  res <- results(dds)

## ----simpleLRT2, eval=FALSE----------------------------------------------
#  dds <- DESeq(dds, test="LRT", reduced=~batch)
#  res <- results(dds)

## ----apeThresh-----------------------------------------------------------
resApeT <- lfcShrink(dds, coef=2, type="apeglm", lfcThreshold=1)
plotMA(resApeT, ylim=c(-3,3), cex=.8)
abline(h=c(-1,1), col="dodgerblue", lwd=2)

## ------------------------------------------------------------------------
condition <- factor(rep(c("A","B","C"),each=2))
model.matrix(~ condition)
# to compare C vs B, make B the reference level,
# and select the last coefficient
condition <- relevel(condition, "B")
model.matrix(~ condition)

## ------------------------------------------------------------------------
grp <- factor(rep(1:3,each=4))
cnd <- factor(rep(rep(c("A","B"),each=2),3))
model.matrix(~ grp + cnd + grp:cnd)
# to compare condition effect in group 3 vs 2,
# make group 2 the reference level,
# and select the last coefficient
grp <- relevel(grp, "2")
model.matrix(~ grp + cnd + grp:cnd)

## ------------------------------------------------------------------------
grp <- factor(rep(1:2,each=4))
ind <- factor(rep(rep(1:2,each=2),2))
cnd <- factor(rep(c("A","B"),4))
model.matrix(~grp + grp:ind + grp:cnd)
# to compare condition effect across group,
# add a main effect for 'cnd',
# and select the last coefficient
model.matrix(~grp + cnd + grp:ind + grp:cnd)

## ----boxplotCooks--------------------------------------------------------
par(mar=c(8,5,2,2))
boxplot(log10(assays(dds)[["cooks"]]), range=0, las=2)

## ----dispFit-------------------------------------------------------------
plotDispEsts(dds)

## ----dispFitCustom-------------------------------------------------------
ddsCustom <- dds
useForMedian <- mcols(ddsCustom)$dispGeneEst > 1e-7
medianDisp <- median(mcols(ddsCustom)$dispGeneEst[useForMedian],
                     na.rm=TRUE)
dispersionFunction(ddsCustom) <- function(mu) medianDisp
ddsCustom <- estimateDispersionsMAP(ddsCustom)

## ----filtByMean----------------------------------------------------------
metadata(res)$alpha
metadata(res)$filterThreshold
plot(metadata(res)$filterNumRej, 
     type="b", ylab="number of rejections",
     xlab="quantiles of filter")
lines(metadata(res)$lo.fit, col="red")
abline(v=metadata(res)$filterTheta)

## ----noFilt--------------------------------------------------------------
resNoFilt <- results(dds, independentFiltering=FALSE)
addmargins(table(filtering=(res$padj < .1),
                 noFiltering=(resNoFilt$padj < .1)))

## ----lfcThresh-----------------------------------------------------------
par(mfrow=c(2,2),mar=c(2,2,1,1))
ylim <- c(-2.5,2.5)
resGA <- results(dds, lfcThreshold=.5, altHypothesis="greaterAbs")
resLA <- results(dds, lfcThreshold=.5, altHypothesis="lessAbs")
resG <- results(dds, lfcThreshold=.5, altHypothesis="greater")
resL <- results(dds, lfcThreshold=.5, altHypothesis="less")
drawLines <- function() abline(h=c(-.5,.5),col="dodgerblue",lwd=2)
plotMA(resGA, ylim=ylim); drawLines()
plotMA(resLA, ylim=ylim); drawLines()
plotMA(resG, ylim=ylim); drawLines()
plotMA(resL, ylim=ylim); drawLines()

## ----mcols---------------------------------------------------------------
mcols(dds,use.names=TRUE)[1:4,1:4]
substr(names(mcols(dds)),1,10) 
mcols(mcols(dds), use.names=TRUE)[1:4,]

## ----muAndCooks----------------------------------------------------------
head(assays(dds)[["mu"]])
head(assays(dds)[["cooks"]])

## ----dispersions---------------------------------------------------------
head(dispersions(dds))
head(mcols(dds)$dispersion)

## ----sizefactors---------------------------------------------------------
sizeFactors(dds)

## ----coef----------------------------------------------------------------
head(coef(dds))

## ----betaPriorVar--------------------------------------------------------
attr(dds, "betaPriorVar")

## ----priorInfo-----------------------------------------------------------
priorInfo(resLFC)
priorInfo(resNorm)
priorInfo(resAsh)

## ----dispPriorVar--------------------------------------------------------
dispersionFunction(dds)
attr(dispersionFunction(dds), "dispPriorVar")

## ----versionNum----------------------------------------------------------
metadata(dds)[["version"]]

## ----normFactors, eval=FALSE---------------------------------------------
#  normFactors <- normFactors / exp(rowMeans(log(normFactors)))
#  normalizationFactors(dds) <- normFactors

## ----offsetTransform, eval=FALSE-----------------------------------------
#  cqnOffset <- cqnObject$glm.offset
#  cqnNormFactors <- exp(cqnOffset)
#  EDASeqNormFactors <- exp(-1 * EDASeqOffset)

## ----lineardep, echo=FALSE-----------------------------------------------
DataFrame(batch=factor(c(1,1,2,2)), condition=factor(c("A","A","B","B")))

## ----lineardep2, echo=FALSE----------------------------------------------
DataFrame(batch=factor(c(1,1,1,1,2,2)), condition=factor(c("A","A","B","B","C","C")))

## ----lineardep3, echo=FALSE----------------------------------------------
DataFrame(batch=factor(c(1,1,1,2,2,2)), condition=factor(c("A","B","C","A","B","C")))

## ----groupeffect---------------------------------------------------------
coldata <- DataFrame(grp=factor(rep(c("X","Y"),each=6)),
                       ind=factor(rep(1:6,each=2)),
                      cnd=factor(rep(c("A","B"),6)))
coldata

## ------------------------------------------------------------------------
as.data.frame(coldata)

## ----groupeffect2--------------------------------------------------------
coldata$ind.n <- factor(rep(rep(1:3,each=2),2))
as.data.frame(coldata)

## ----groupeffect3--------------------------------------------------------
model.matrix(~ grp + grp:ind.n + grp:cnd, coldata)

## ----groupeffect4, eval=FALSE--------------------------------------------
#  results(dds, contrast=list("grpY.cndB","grpX.cndB"))

## ----missingcombo--------------------------------------------------------
group <- factor(rep(1:3,each=6))
condition <- factor(rep(rep(c("A","B","C"),each=2),3))
d <- DataFrame(group, condition)[-c(17,18),]
as.data.frame(d)

## ----missingcombo2-------------------------------------------------------
m1 <- model.matrix(~ condition*group, d)
colnames(m1)
unname(m1)
all.zero <- apply(m1, 2, function(x) all(x==0))
all.zero

## ----missingcombo3-------------------------------------------------------
idx <- which(all.zero)
m1 <- m1[,-idx]
unname(m1)

## ----cooksPlot-----------------------------------------------------------
W <- res$stat
maxCooks <- apply(assays(dds)[["cooks"]],1,max)
idx <- !is.na(W)
plot(rank(W[idx]), maxCooks[idx], xlab="rank of Wald statistic", 
     ylab="maximum Cook's distance per gene",
     ylim=c(0,5), cex=.4, col=rgb(0,0,0,.3))
m <- ncol(dds)
p <- 3
abline(h=qf(.99, p, m - p))

## ----indFilt-------------------------------------------------------------
plot(res$baseMean+1, -log10(res$pvalue),
     log="x", xlab="mean of normalized counts",
     ylab=expression(-log[10](pvalue)),
     ylim=c(0,30),
     cex=.4, col=rgb(0,0,0,.3))

## ----histindepfilt-------------------------------------------------------
use <- res$baseMean > metadata(res)$filterThreshold
h1 <- hist(res$pvalue[!use], breaks=0:50/50, plot=FALSE)
h2 <- hist(res$pvalue[use], breaks=0:50/50, plot=FALSE)
colori <- c(`do not pass`="khaki", `pass`="powderblue")

## ----fighistindepfilt----------------------------------------------------
barplot(height = rbind(h1$counts, h2$counts), beside = FALSE,
        col = colori, space = 0, main = "", ylab="frequency")
text(x = c(0, length(h1$counts)), y = 0, label = paste(c(0,1)),
     adj = c(0.5,1.7), xpd=NA)
legend("topright", fill=rev(colori), legend=rev(names(colori)))

## ----vanillaDESeq, eval=FALSE--------------------------------------------
#  dds <- DESeq(dds, minReplicatesForReplace=Inf)
#  res <- results(dds, cooksCutoff=FALSE, independentFiltering=FALSE)

## ----varGroup, echo=FALSE------------------------------------------------
set.seed(3)
dds1 <- makeExampleDESeqDataSet(n=1000,m=12,betaSD=.3,dispMeanRel=function(x) 0.01)
dds2 <- makeExampleDESeqDataSet(n=1000,m=12,
                                betaSD=.3,
                                interceptMean=mcols(dds1)$trueIntercept,
                                interceptSD=0,
                                dispMeanRel=function(x) 0.2)
dds2 <- dds2[,7:12]
dds2$condition <- rep("C",6)
mcols(dds2) <- NULL
dds12 <- cbind(dds1, dds2)
rld <- rlog(dds12, blind=FALSE, fitType="mean")
plotPCA(rld)

## ----convertNA, eval=FALSE-----------------------------------------------
#  res$padj <- ifelse(is.na(res$padj), 1, res$padj)

## ----sessionInfo---------------------------------------------------------
sessionInfo()


```


```{r, include=FALSE}
library(seqinr)
## Workflow to create senquence logo's in R
# https://davetang.org/muse/2013/01/30/sequence-logos-with-r/ 
library(Biostrings)  
library(seqLogo)

# Create position weight matrix

## Example sequences
sequences <- c("GCCAAAAAAA",
               "CGCAAAAAAA",
               "TCATAAAAAC",
               "GCATTAAAAA",
               "TCGTAAAAAC")
               
#?consensusMatrix

sequences <- DNAStringSet(sequences)
cmx <- consensusMatrix(sequences)
cmx

hasOnlyBaseLetters(sequences)

clean_cmx <- alphabetFrequency(sequences, baseOnly=TRUE)
               
#install if necessary
#source("http://bioconductor.org/biocLite.R")
#biocLite("seqLogo")



#create data frame using the four vectors

#define function that divides the frequency by the row sum i.e. proportions
proportion <- function(x){
  rs <- sum(x);
  return(x / rs);
}



#create position weight matrix
clean_cmx <- t(clean_cmx)
clean_cmx <- as.data.frame(clean_cmx[c(1:4),])
clean_cmx

pwm <- apply(clean_cmx, 2, proportion)
pwm
pwm <- makePWM(pwm)
```

## Sequence logog plot
```{r}
#png("hunchback.png")
seqLogo(pwm)
# dev.off()

```



## Interactive plot
```{r}

library()






```


