---
title: "R basics - Exercise"
author: "Marc A.T. Teunis"
date: '`r Sys.Date()`'
output: 
  learnr::tutorial:
    progressive: true
    allow_skip: true
runtime: shiny_prerendered
---

This document is an R Markdown Tutorial (create with the `learnr` package)
An example of the use of _literate programming_ in R. R Markdown is based on the Markdown mark-up language used for easy writing for the web and can combine plain human-readable text with computer code and output from that code in one document. 

`learnr` is an implementation of `Shiny` apps. See http://www.rstudio.com/shiny for more information.

Type code in empty blocks or run the pre-defined code by clicking the `Run Code` button.
Clicking `Start Over` clears the exercise and enables you to begin again.

## The data
In the first part of this tutorial we use the data from the the in R available demo dataset: `Theoph`

Boeckmann, Sheiner and Beal (1994) report data from a study by Dr. Robert Upton of the kinetics of the anti-asthmatic drug theophylline. Twelve subjects were given oral doses of theophylline then serum concentrations were measured at 11 time points over the next 25 hours.

```{r setup, include=FALSE}
#knitr::opts_chunk$set(echo = TRUE, 
#                      warning = TRUE, 
#                      message = TRUE, 
#                      error = TRUE)
library(learnr)
tutorial_options(exercise.timelimit = 60)
#tutorial_options(exercise.checker = checkthat::check_exercise)
knitr::opts_chunk$set(error = TRUE)
```

## Packages
```{r load_packages, exercise = TRUE}
library(tidyverse)
library(utils)
```

## Aim of the tutorial
In this tutorial you will:

 * Get a first glimpse at R and it's language syntax
 * Write a few lines of code
 * Load an example dataset
 * Create exploratory data visualizations 
 * Inspect thee dataset
 * Do calculations

## Load example dataset

Type `data` in the field below, under the predefined code. Click `Run Code`
What happens?

```{r load_data, exercise = TRUE, exercise.lines = 7, warning=FALSE, error=FALSE}
library(tidyverse)
library(datasets)
data <- as_tibble(datasets::Theoph)




```

## Getting help

To get help on a function, a package or a dataset, type `?package/function/dataset_name`
Can you use `?dataset_name` to learn something about the `` dataset.
```{r getting_help, exercise = TRUE}
?datasets::Theoph

```

From the help on `Theoph` answer the following question.
```{r quiz, echo=FALSE}
quiz(
  question("Theoph is short for theophylline, which is an",
    answer("Anti-inflammatory drug"),
    answer("Anti-asthmatic drug", correct = TRUE),
    answer("Anti-rheumatoid drug"),
    answer("Anti-cancer drug"),
allow_retry = TRUE, random_answer_order = TRUE))
```

## Inspecting a dataframe

Click "Run Code" to see the code

 - How do we inspect `data`?
 - How many variables does `data` have?
 - How many rows?
 - Call `summary()` on `data`; 
 In which variable do we observe the highest measured value ratio?

Try the code in the Code fields or type in the fields and 
click "Run Code"

### Head tail, and dimensions
Try the commands `head(data)`, `tail(data)` and `dim(data)` below.
```{r, inspect_th_a, exercise = TRUE}
data <- as_tibble(datasets::Theoph)
head(data)
class(data)

## use tail() and dim() functions on data:


```

### Number of variables
Type `ncol(data)`. What does this command do?
```{r, inspect_th_b, exercise = TRUE}
data <- as_tibble(datasets::Theoph)


```

### Number of rows
Type `nrow(data)`. What does this do? 
```{r, inspect_th_c, exercise = TRUE}
data <- as_tibble(datasets::Theoph)


```

## Exploring the `Theop` data

R is great for visualizing you data. Discuss with your neighbour: which variables would you plot first in this dataset?

Run the code below to see the first graph
```{r, plot_1, exercise = TRUE}
data <- as_tibble(datasets::Theoph)
head(data)

plot_1 <- data %>%
  ggplot(aes(x = conc)) +
  geom_histogram(aes(y=..density..),      
                   bins = 30,
                   colour = "black", fill = "white") +
    geom_density(alpha = .2, fill = "Blue") + xlab("value") +
  xlab("Concentration")

plot_1

```

### Scatter plots are a good way to start too
What do you think about the relationship btween Time and concentration of theophylline?
```{r, plot_2_scatter, exercise = TRUE}
data <- as_tibble(datasets::Theoph)
plot_2 <- data %>%
  ggplot(aes(x = Time, y = conc)) +
  geom_point() 

plot_2
```

### Adding a smooth line can help discover a pattern.
What do you think is the grey area? 

```{r, plot_3_smooth, exercise = TRUE}
data <- as_tibble(datasets::Theoph)

plot_3 <- data %>%
  ggplot(aes(x = Time, y = conc)) +
  geom_point()  + 
  geom_smooth()

plot_3
```

### Different doses 
Different doses were administered. Discuss with you neighbour how you could visualize this.
Let's add a bit of colour. What do you think is the realtionship between Dose and Subject. Did every Subject recieve the same dose?
```{r, plot_4_colours, exercise = TRUE}
data <- as_tibble(datasets::Theoph)

data$Dose <- as.factor(data$Dose)

plot_4 <- data %>%
  ggplot(aes(x = Time, y = conc)) +
  geom_point(aes(colour = Dose)) 

plot_4
```

### Where are the individual Subjects in the data?
Run the code below. what does the command `facet_wrap(~ Subject) do?
What do you think, are there many differences between subjects in the kinectics of theophylline in vivo?
```{r, plot_5_facets, exercise = TRUE}
data <- as_tibble(datasets::Theoph)

data$Dose <- as.factor(data$Dose)

plot_5 <- data %>%
  ggplot(aes(x = Time, y = conc)) +
  geom_point(aes(colour = Dose)) +
  facet_wrap(~ Subject)

plot_5
```

### Adding a line per subject is easy!
```{r, plot_6_smoothers, exercise = TRUE}
data <- as_tibble(datasets::Theoph)

data$Dose <- as.factor(data$Dose)

plot_6 <- data %>%
  ggplot(aes(x = Time, y = conc)) +
  geom_point(aes(colour = Dose)) +
  geom_smooth(se = FALSE) +
  facet_wrap(~ Subject)

plot_6
```


## Summary data
Call `summary(Theoph)`. Can you determine the maximum value in the dataset?
How can you use the command `max()` to achieve the same?
```{r, inspect_th_d, exercise = TRUE}



```

```{r, inspect_th_d-solution, include=FALSE}
summary(Theoph)


```

```{r, inspect_th_d-check}

"Well done!"

```

## Another example; Toxicity data

From Teunis M., Corsini E., Smits M, Madsen C.B., Eltze T., Ezendam J., Galbiati V.,
Gremmer E., Krul C., Landin A., Landsiedel R., Pieters R., Rasmussen T.F., Reinders J.,
Roggen E., Spiekstra S., Gibbs S. Transfer of a two-tiered keratinocyte assay:
IL-18 production by NCTC2544 to determine the skin sensitizing capacity and
epidermal equivalent assay to determine sensitizer potency. Toxicol In Vitro.
2013 Apr;27(3):1135-50. doi: 10.1016/j.tiv.2012.06.004. Epub 2012 Jun 19. PubMed 
PMID: 22728231.

http://www.sciencedirect.com/science/article/pii/S0887233312001543?via%3Dihub 

Run the Code below. Do you remember from above how to inspect the data to learn what variables it contains?

### Loading data

<div id="create_vector_b-hint">
**Hint:** To inspect the data call `head(data_table)`.
</div>

```{r, nctc_load, exercise = TRUE}
library(readxl)
library(readr)

URL <- "https://www.innovativetesting.nl/sites/default/files/attachment/NCTC%20data%20all%20participants%20DTU%20and%20UNIMI%20corrected%20%282%29.csv"
destfile <- "nctc.csv"
download.file(URL, destfile)

nctc <- read_csv(file = destfile)

nctc <- nctc[-c(1:2),c(1:8)]
nctc 
```

### Cleaning the variables in the dataset
The names of the variables in the `nctc` datatable can be viewed by calling `names(nctc)`
```{r, nctc_clean, exercise = TRUE}

library(tidyverse)
URL <- "https://www.innovativetesting.nl/sites/default/files/attachment/nctc2.csv"
destfile <- "nctc.csv"
download.file(URL, destfile)

#nctc <- readRDS(file = "./nctc.RDS")
 nctc <- read_csv(file = destfile)
nctc <- nctc[-c(1:2),c(1:8)]

names(nctc) <- tolower(names(nctc))
names(nctc) <- gsub(pattern = "-", replacement = "", names(nctc))
names(nctc) <- gsub(pattern = "\\.", replacement = "", names(nctc))

nctc$lab <- as.factor(nctc$lab)
nctc$compound <- as.factor(nctc$compound)
nctc$conc <- as.factor(nctc$conc)
nctc$exp <- as.factor(nctc$exp)
nctc$rep <- as.factor(nctc$rep)
nctc$viability <- as.numeric(nctc$viability)
nctc$il18 <- as.numeric(nctc$fold)

head(nctc)

## saveRDS(nctc, file = "./nctc.RDS")

```

## Exploratory data analysis; NCTC data
```{r, eda_nctc, exercise = TRUE}
#nctc <- readRDS(file = "./nctc.RDS")
nctc <- read_csv(file = destfile)
nctc <- nctc[-c(1:2),c(1:8)]

names(nctc) <- tolower(names(nctc))
names(nctc) <- gsub(pattern = "-", replacement = "", names(nctc))
names(nctc) <- gsub(pattern = "\\.", replacement = "", names(nctc))

nctc$lab <- as.factor(nctc$lab)
nctc$compound <- as.factor(nctc$compound)
nctc$conc <- as.factor(nctc$conc)
nctc$exp <- as.factor(nctc$exp)
nctc$rep <- as.factor(nctc$rep)
nctc$viability <- as.numeric(nctc$viability)
nctc$il18 <- as.numeric(nctc$fold)


nctc_summary <- nctc %>%
  na.omit() %>% 
  group_by(conc, compound) %>%
  summarise(mean_viab = mean(viability),
            mean_il18 = mean(il18),
            mean_fold = mean(fold))
head(nctc_summary)  
summary(nctc_summary)

saveRDS(nctc_summary, file = "./nctc_summary.RDS")

```

### All compounds
```{r, all_compounds, exercise = TRUE}
library(tidyverse)

nctc_summary <- readRDS(file = "./nctc_summary.RDS")

ggplot(data = nctc_summary, aes(x = compound, y = mean_viab)) +
  geom_point() +
  geom_hline(aes(yintercept = 100), colour = "Blue", size = 2)

```

### One compound
Let's select the compound with the highest IL-18 induction
Which compund has highest IL-18 induction?
```{r, highest_il-18, exercise = TRUE}
library(tidyverse)
nctc_summary <- readRDS(file = "./nctc_summary.RDS")

high_il18 <- nctc_summary %>% 
 filter(mean_il18 >= 3.0) %>%
  arrange(desc(mean_il18))

```

### Highest IL-18 plot
```{r, highest_il-18_plot, exercise = TRUE}
library(tidyverse)

nctc_summary <- readRDS(file = "./nctc_summary.RDS")

compound_x <- nctc_summary %>%
  filter(compound == "58") %>%
  arrange(conc)
 
ggplot(data = compound_x, aes(x = conc, y = mean_il18)) +
  geom_bar(position = "dodge", stat = "identity")
```

### Let's add a compound that has no high IL-18 expression
```{r, add_no_high_il-18_to_plot, exercise = TRUE}
library(tidyverse)
nctc_summary <- readRDS(file = "./nctc_summary.RDS")

low_il18 <- nctc_summary %>% 
 filter(mean_il18 <= 1.0) %>%
  arrange(desc(mean_il18))

compound_xy <- nctc_summary %>%
  filter(compound == "58" | compound == "121") %>%
  arrange(compound, conc)
 
ggplot(data = compound_xy, aes(x = conc, y = mean_il18)) +
  geom_bar(aes(fill = compound), position = "dodge", stat = "identity")

```

